---
title: "U.S. Weather Activities versus the Public Health"
author: "Jim Stephenson"
date: "August 24, 2014"
output:
  html_document:
    fig_caption: yes
    highlight: tango
    keep_md: yes
---

## U.S. Weather Activities versus the Public Health
### Synopsis

Data for this project comes from NOAA's storm database.  The [data file](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) is a fairly large (47MB) bzipped file that decompresses to a large (562MB) CSV file.  Within the data file are records of weather events spanning 1950 to 2011.

```{r loadData, echo=TRUE, cache=TRUE, message=FALSE}
## download and extract the data 
if (!file.exists("./data")) {
  dir.create("./data")
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileURL, destfile = "./data/dataset.bz2", method = "curl", mode = "wb")
}
# note: this next line will take a *very* long time to execute (5 or more minutes)
# ... go make a cup of coffee ...
storm <- read.csv(bzfile("./data/dataset.bz2", "r"))
```
After reading in the data, we can see that there are over 900K records, with 37 variables.  Fortunately, the columns we care about -- Event Type (EVTYPE), Fatalities (FATALITIES), and Injuries (INJURIES) -- do not contain NA values, so we don't have to manage them:

```{r displayData, echo=TRUE, cache=TRUE, message=FALSE}
dim(storm)
na.test <-c(anyNA(storm$EVTYPE),anyNA(storm$FATALITIES), anyNA(storm$INJURIES))
na.test
```
### Loading and Processing Raw Data



### Results